Tracing the cultural evolution of meaning at the semantics-pragmatics interface

### 2016-11-04 paper/paper-v2.tex
minor changes in draft

\title{Communicative pressures at the semantics-pragmatics interface}

@@

\begin{abstract}
According to standard linguistic theory, the conventional meaning of expressions is systematically modified by general pragmatic rules governing their use. This implies that models of the cultural evolution of meaning should likewise take into consideration that observable language use is a complex interaction of semantic representations and pragmatic rules. To this end, we present a game theoretic model of the cultural evolution of language where communicative pressures work on abstract semantic representations and pragmatic usage rules. Our model traces two evolutionary forces and their interaction: (i) fitness-based pressure towards communicative efficiency and (ii) systematic transmission perturbations when linguistic knowledge is transferred from one agent to another. The latter can contain general cognitive or learning biases, but can also contain other sources of systematic noise, e.g., resulting from errors in perception. We illustrate the model based on a case study showing that cognitive biases that favor simple semantic representations can prevent the lexicalization of pragmatic inferences. We also show that, technically speaking, it is possible that environmental factors, such as perceptual errors during acquisition, can produce evolutionary outcomes that look as if such cognitive biases are present even if they are not. 
\end{abstract}

%Mutual reasoning endows speakers with the ability to convey information that goes beyond the literal meaning of expressions. A particularity of certain classes of lexical meanings is that they enable for such pragmatic enrichments in a notably productive fashion. This raises the challenge to justify their regular selection over other alternatives that, e.g., do not require such enrichments. We suggest that such an account can be provided by an analysis of the linguistic pressures that apply at the semantics-pragmatics interface. To this end, we lay out a general model that combines functional pressure on language use and a pressure for learnability in acquisition in populations of probabilistic language users with varied degrees of pragmatic sophistication. We then apply this model in a case study on the (lack of) lexicalization of scalar implicatures. The results suggest that such inferences fail to lexicalize when languages are pressured towards simpler semantic representations in learning, provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages otherwise incur. This result sheds light on the lack of lexicalization of scalar inferences, as well as the semantics-pragmatics distinction more generally.

\section{Introduction}\label{sec:introduction}
What is conveyed usually goes beyond what is said. A request for a blanket can be politely veiled by uttering ``I'm cold'', a temporal succession of events can be conveyed by the order in which conjuncts appear as in ``I traveled to Paris and got married'', or an invitation can be declined by saying ``I have to work''. An influential explanation of the relation between the literal meaning of expressions (semantics) and what they are intended and interpreted to convey (pragmatics) due to \citet{grice:1975} views the latter as a product of a process of mutual reasoning about rational language use. For instance, under the assumption that the speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a reason why the speaker will not be able to accept the invitation -- why would she have said so otherwise? Crucially, many classes of lexical meanings allow for pragmatic enrichments in a notably systematic fashion.

Productive pragmatic inferences such as scalar and manner implicatures have been at the focus of many studies on the semantics-pragmatics interface. However, an issue that has received little attention is the justification of semantic structure in light of the systematic enrichments offered by pragmatics. Similarly, multiple models of the cultural evolution of language have been put forward to analyze properties such as compositionality and combinatoriality, but the selection of particular semantics in light of pragmatics remains uncharted territory. The present investigation seeks to fill these gaps by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of their systematic pragmatic modification. Accordingly, our main goal is to uncover how transmission biases and functional pressure interact in the cultural evolution of meaning when pragmatics is taken into account. Crucially, we argue that transmission biases need not be cognitive in nature but can also ensue from perception and the environment.

\subsection{Linguistic pressures at the semantics-pragmatics interface}
The emergence and change of linguistic structure is influenced by many intertwined factors. These range from biological and socio-ecological to cultural ones \citep{steels:2011}. Social and ecological pressures determine communicative needs while biology determines the architecture that enables and constrains the means by which they can be fulfilled. In the following, our focus lies on cultural aspects, wherein processes of linguistic change are viewed as shaped by language use and its transmission. That is, as a result of a process of cultural evolution.

The idea that language is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses at latest since \citeposs{zipf:1949} rationalization of the approximation of word frequency rankings by a power law distribution as competing hearer and speaker preferences (e.g. in \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}). In recent years this line of research has led to a surge of approaches that seek to analyze the effects of such pressures from multiple angles ranging from simulations to experiments with human and robotic agents (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews). Our starting point is given by the overarching argument that has crystalized from this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations. More succinctly; natural languages are pressured for expressivity and learnability.

The opposition of expressivity and learnability becomes particularly clear when considering their consequences in the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a single form-meaning association is easy to learn but lacking in expressivity. Conversely, a language that associates a distinct form with all possible meanings one may want to convey is maximally expressive but challenging to acquire. The most prominent problem that arises from this tension is that of acquiring a language to express a potentially infinite set of meanings through finite means \citep{kirby:2002}. However, this is not the only challenge learners confront. More central to our explanandum is the issue of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to. At the semantics-pragmatics interface this concerns the selection between functionally (near-)equivalent lexical meanings, noting in particular that what is systematically conveyed through pragmatics could alternatively be codified lexically. Notwithstanding, lexical meanings that allow for pragmatic modifications often seem to resist the lexicalization of their pragmatic component. The question is why.

We assume an integral part of the answer to lie in the effects of transmission perturbations on language when passed from one generation to the next. Such perturbations may take the form of learning biases, but may also stem from extraneous factors such as a noisy perception of the environment. In particular, we show how the presence of systematic noise can introduce regularizations that resemble those obtained under the assumption of a learning bias.

In the following, we model these components using the replicator-mutator dynamics, combining functional pressure on successful communication with effects of transmission perturbations on (iterated) Bayesian learning \citep{griffiths+kalish:2007}. The semantics-pragmatics distinction and its bearing on production and comprehension are captured by a probabilistic model of rational language use with different degrees of pragmatic sophisitication and languages \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}. The remainder of this section introduces these components individually together with the assumptions underlying them. These are: the representation of languages and their use (\S\ref{sec:languages+use}), pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}) regulated by the replicator and mutator dynamics, respectively. After laying out the model, we analyze its predictions in a case-study on the lack of lexicalization of scalar implicatures in \S\ref{sec:si-case-study} and compare the outcome predicted by different transmission biases in \S\ref{subsec:bias} and \S\ref{subsec:noise}.

\subsection{Lexica and linguistic behavior}\label{sec:languages+use}
Lexica codify the truth-conditions of a language's expressions. Following \citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) to convey and $M$ a set of messages (forms available in the language).

We distinguish between two kinds of linguistic behavior. {\em Literal interlocutors} produce and interpret messages literally, being guided only by their lexica. In contrast, {\em pragmatic interlocutors} engage in mutual reasoning to inform their choices. Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic predecessors, the Iterated Best/Quantal Response models \citep{franke:2009,franke+jaeger:2014}, these behaviors are captured by a reasoning hierarchy. The hierarchy's bottom, level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$ behave rationally according to expected level $n$ behavior of their interlocutors. Below, (\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and pragmatic hearers of a language $L$. Their speaker counterparts are given in (\ref{s:level0}) and (\ref{s:leveln}).

\begin{flalign}
&H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m|s;L) \propto \exp(L_{sm}^\alpha) \label{s:level0}\\
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)^\alpha) \label{s:leveln}
\end{flalign}

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. In the following this prior is assumed to be uniform across hearers. The behavior of literal speakers, given in (\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the sensitivity to which speakers prefer one signal over another based on its expected communicative success.

@@

We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a {\em type}. These are the basic units on which our model's dynamics operate.

@@

The replicator equation gives us the means to make these dynamics precise. The proportion of types in a given population is codified in a vector $x$, where $x_i$ is type $i$'s proportion. As noted above, the fitness of type $i$ is equal to its relative communicative success within this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$. The expected communicative success of $i$ and $j$ is obtained by considering the average success of $i$ conveying information to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$. $U_S(x,y)$ and $U_R(x,y)$ are, respectively, $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$, for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise. This quantity is symmetric, reflecting the probability of two types' mutual understanding. The average fitness of the population is given by $\Phi$, $\Phi = \sum_i x_i f_i$. This term serves as a normalizing constant for the discrete replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$

Under its biological interpretation, this equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations. As noted above, many aspects of language are subject to processees of transmission and change that can be likened to these biological dynamics. Amongst others, replication can be construed as modelling language acquisition, as e.g. in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation in a single generation (see \citealt[\S3.3]{benz+etal:2005b} for discussion).

In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role of the fidelity by which languages are transmitted. Crucially, linguistic production can be prone to errors, states or messages may be percieved incorrectly, and multiple languages may be compatible with the data learners are exposed to. These sources of uncertainty introduce variation in their transmission from one generation to the next. In keeping the analogy to evolutionary processes this variation can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent. Importantly, the resulting generational turnovers should depend on the relative learnability of a type instead of being a constant that is equal to all types as in \citet{nowak+etal:2002}. For this purpose, we turn to a different strand of research in cultural evolution: {\em iterated learning}.

@@

\subsection{Learnability}\label{sec:learnability}
Iterated learning is a process in which the behavior of one individual serves as learning input for another. This learner, upon acquisition of this behavior, then goes on to produce behavior that serves as input for a new learner. This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language. The latter, now a parent, goes on to produce linguistic data for a new generation of learners. Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated process of Bayesian inference in which learners combine the likelihood of a type producing the received learning input with prior inductive biases.

Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for recent surveys). Importantly, experimental and mathematical results suggest that the outcome of this process reflects learners' a priori biases. In a Bayesian setting these biases can be codified in a prior $P \in \Delta(T)$ and a way to think about it is as the amount of data a learner would require in order to adopt a language (cf. \citealt[450]{griffiths+kalish:2007}). Or, in our case, a combination of a lexicon and a signaling behavior. The extent of the prior's influence has been shown to heavily depend on the learning strategy assumed to underly the inference process. On the one hand, simulation results suggested that weak biases could be magnified by exposing learners to only small data samples (e.g. in \citealt{brighton:2002}). On the other, \citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated learning converged to the prior, i.e., the resulting distribution over languages corresponds to the learners' prior distribution and is not influenced by the amount of input given to them. This difference in predictions can be traced back to differences in the selection of hypotheses from the posterior. Griffith \& Kalish's convergence to the prior holds for learners that sample from the posterior. More deterministic strategies such as the adoption of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we parametrize the posterior with a parameter $l \geq 1$. In doing so, we obtain a range of learning strategies between posterior sampling and MAP. When $l = 1$ learners sample from the posterior. As $l$ increases towards infinity their propensity to maximize the posterior grows as well.

@@

%To summarize, the parametrized posterior, $P(d|t)^l$, is obtained from combining the prior $P(t)$ and the likelihood $P(d|t)$. Since types are combinations of a lexicon and its use, the latter can be computed straightforwardly from a type's production behavior. 
We integrated iterated learning in the replicator dynamics by codifying the former in a transition matrix $Q$, where $Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$. This quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data:

@@

\subsection{Summary}
Drawing from past research, we argued that expressivity and learnability are central to the cultural evolution of language. We propose these components to be modelled respectively as communicative efficiency-relative replication and (iterated) Bayesian learning. Taken together their interaction is described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:

@@

\section{Scalar implicatures}\label{sec:si-case-study}
%
Scalar implicatures are a particularly well-studied type of conventional pragmatic inferences. They are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order. For instance, {\em some} is entailed by {\em all}. If it were true that `All students came to class', it would also be true that `Some students came to class'. However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey. Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold because the speaker used a weaker alternative instead. In this way, `Some students came to class' is strengthened to convey that some but not all students came to class. Conversely, speakers can rely on their interlocutors to draw this inference. A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.

This kind of strengthening is captured by the signaling behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about a speaker's use of message involving {\em some} will associate it more strongly with a state in which {\em some but not all} holds than with one in which the stronger alternative {\em all} holds, as the {\em all} already unambiguously conveys the latter state. Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly.

@@

We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions. The first is that their truth-conditional compatibility with stronger expressions endows them with a broader range of applicability by allowing them to occur in contexts in which their upper-bounded reading is absent. This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant. For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty. This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions. For example, by saying {\em some but not all} explicitly. Importantly, although morphosyntactic disambiguation may be dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer. In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so. Compare this with a hypothetical language that lexicalizes two expressions for each weak scalar expression -- one with and one lacking an upper-bound. We see four conditions along this functionalist explanation that may pressure languages for English-like semantics over this alternative. First, contextual cues are very reliable. Second, morphosyntactic disambiguation is seldom necessary. Third, morphosyntactic disambiguation is only marginally dispreferred. Fourth, larger lexica are costly. Overall, neither condition is convincing as a pivotal explanatory device for such a wide-spread phenomenon. The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication. It is likely that human language users are very good at retrieving cues from context, but to stipulate that they are so good as to undercut the benefit of safe communication provided by this hypothetical alternative strikes us as too strong of an assumption.  As for the third and fourth condition, these seem mostly like technical solutions without a proper empirical basis.

>>>>

\title{Tracing the cultural evolution of meaning at the semantics-pragmatics interface}

@@

\begin{abstract}
According to standard linguistic theory, the conventional meaning of expressions is systematically modified by general pragmatic rules governing their use. This implies that models of the cultural evolution of meaning should likewise take into consideration that observable language use is a complex interaction of semantic representations and pragmatic rules. To this end, we present a game theoretic model of the cultural evolution of language where communicative pressures work on abstract semantic representations and pragmatic usage rules. Our model traces two evolutionary forces and their interaction: (i) fitness-based pressure towards communicative efficiency and (ii) systematic transmission perturbations when linguistic knowledge is transferred from one agent to another. The latter can arise from general cognitive or learning biases, but also from other sources of systematic noise, e.g., as a result of errors in perception. We illustrate the model based on a case study showing that cognitive biases that favor simple semantic representations can prevent the lexicalization of pragmatic inferences. We also show that, technically speaking, it is possible that environmental factors, such as perceptual errors during acquisition, can produce evolutionary outcomes that look as if such cognitive biases are present even if they are not. 
\end{abstract}

\section{Introduction}\label{sec:introduction}
What is conveyed usually goes beyond what is said. A request for a blanket can be politely veiled by uttering ``I'm cold'', the temporal succession of events can be conveyed by the order in which conjuncts appear as in ``I traveled to Paris and got married'', or an invitation be declined by saying ``I have to work''. An influential explanation of the relation between the literal meaning of expressions and what they are intended and interpreted to convey due to \citet{grice:1975} views the latter as a product of a process of mutual reasoning about rational language use. For instance, under the assumption that the speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a reason why the speaker will not be able to accept the invitation, going beyond its literal meaning. Crucially, many classes of lexical meanings allow for pragmatic enrichments in a notably systematic fashion.

Productive pragmatic inferences, such as scalar and manner implicatures, have been at the focus of many studies on the semantics-pragmatics interface. However, an issue that has received little attention is the justification of semantic structure in light of the systematic enrichments offered by pragmatics. Similarly, differnt models of the cultural evolution of language have been put forward in order to analyze properties such as compositionality and combinatoriality (e.g. \citealt{kirby+etal:2015,verhoef+etal:2014}), but the selection of meaning in light of pragmatics remains uncharted territory. The present investigation seeks to fill these gaps by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of their pragmatic modification. Accordingly, our primary goal is to uncover how transmission perturbations and functional pressure interact in the cultural evolution of meaning. Additionally, we show how systematic disturbances in the fidelity by which linguistic knowledge is transmitted that stem from environmental factors can lead to outcomes akin to those predicted under the assumption of inductive learning biases. That is, that the regularization standardly predicted to follow from learning biases can also emerge as a byproduct of noise.

\subsection{Communicative pressures at the semantics-pragmatics interface}
The emergence and change of linguistic structure is influenced by many intertwined factors. These range from biological and socio-ecological to cultural ones \citep{steels:2011}. Social and ecological pressures determine communicative needs, while biology determines the architecture that enables and constrains the means by which they can be fulfilled. In the following, our focus lies on the cultural aspects, wherein processes of linguistic change are viewed as shaped by language use and its transmission, i.e., as a result of a process of cultural evolution.

The idea that language is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses at latest since \citeposs{zipf:1949} rationalization of the approximation of word frequency rankings by a power law distribution as competing hearer and speaker preferences (e.g. in \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}). In recent years, this line of research has led to a surge of approaches that seek to analyze the effects of such pressures from a multidisciplinary perspective, ranging from simulations to experiments with human and robotic agents (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews). Our starting point is given by the overarching argument that has crystalized from this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations. More succinctly; natural languages are pressured for expressivity and learnability.

The opposition of expressivity and learnability becomes particularly clear when considering their consequences in the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a single form-meaning association is easy to learn but lacking in expressivity. Conversely, a language that associates a distinct form with all possible meanings a speaker may want to convey is maximally expressive but challenging to acquire. The most prominent problem that arises from this tension is that of acquiring a language to express a potentially infinite set of meanings through finite means \citep{kirby:2002}. However, this is not the only challenge learners confront. More central to our explanandum is the issue of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to. At the semantics-pragmatics interface this concerns the selection between functional (near-)equivalents, noting in particular that what is systematically conveyed through pragmatics could alternatively be codified lexically. Notwithstanding, lexical meanings that allow for pragmatic modifications often seem to resist the lexicalization of their pragmatic component. The question is why.

We assume an integral part of the answer to lie in the effects of transmission perturbations that arise when linguistic knowledge is acquired by new learners. As shown in the following, such perturbations may take the form of learning biases, but may also stem from extraneous factors such as a noisy perception of the environment.

We model these components using the replicator-mutator dynamics, combining functional pressure on successful communication with effects of transmission perturbations on (iterated) Bayesian learning \citep{griffiths+kalish:2007}. The semantics-pragmatics distinction and its bearing on production and comprehension are captured by a probabilistic model of rational language use with different degrees of pragmatic sophisitication and different lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}. The remainder of this section introduces these components together with the assumptions underlying them. These are: the representation of languages and their use (\S\ref{sec:languages+use}), as well as pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}). After laying out the model, we analyze its predictions in a case-study on the lack of lexicalization of scalar implicatures in \S\ref{sec:si-case-study}. Additionally, \S\ref{subsec:noise} shows how the outcome predicted under the assumption of an inductive bias for simplicity in \S\ref{subsec:bias} can also arise from as a byproduct of the noisy perception of language users and learners. Section \S\ref{sec:discussion} discusses the general predictions of the model, as well as the implications drawn from its noisy counterpart.

\subsection{Lexica and linguistic behavior}\label{sec:languages+use}
Lexica codify the truth-conditions of a language's expressions. Following \citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) and $M$ a set of messages (forms available in the language).

We distinguish between two kinds of linguistic behavior. {\em Literal interlocutors} produce and interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors} instead engage in mutual reasoning to inform their choices. Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic predecessors, the Iterated Best/Quantal Response models \citep{franke:2009,franke+jaeger:2014}, these behaviors are captured by a reasoning hierarchy. The hierarchy's bottom, level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$ behave rationally according to expected level $n$ behavior of their interlocutors. (\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and pragmatic hearers of a language $L$. Their speaker counterparts are given in (\ref{s:level0}) and (\ref{s:leveln}).

\begin{flalign}
&H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha) \label{s:level0}\\
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)^\alpha) \label{s:leveln}
\end{flalign}

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in the following this prior is assumed to be uniform across hearers. The behavior of literal speakers, given in (\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the sensitivity to which speakers prefer one signal over another based on its expected communicative success.

@@

We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a {\em type}. These are the basic units on which the model's dynamics operate.

@@

The replicator equation gives us the means to make these dynamics precise. The proportion of types in a given population is codified in a vector $x$, where $x_i$ is type $i$'s proportion. As noted above, the fitness of type $i$ is equal to its relative communicative success within this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$. The expected communicative success of $i$ and $j$ is simply the average success of $i$ conveying information to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$. $U_S(x,y)$ and $U_R(x,y)$ are, respectively, $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$, for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise. This quantity is symmetric, reflecting the probability of two types' mutual understanding. The average fitness of the population is given by $\Phi$, $\Phi = \sum_i x_i f_i$, serving as a normalizing constant for the discrete replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$

Under its biological interpretation, the replicator dynamics capture the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations. Many aspects of language are subject to transmission and change that can be likened to such biological processes. Amongst others, replication can be construed as modelling language acquisition, as e.g. in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation in a single generation (see \citealt[\S3.3]{benz+etal:2005b} for discussion).

In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role of the fidelity by which language is transmitted. Among others, linguistic production can be prone to errors, states or messages may be percieved incorrectly, and multiple languages may be compatible with the data learners are exposed to. These sources of uncertainty introduce variation in their transmission from one generation to the next which can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent as a result of transmission perturbations. Importantly, this variation should be relative to the learnability of a type under these perturbations (instead of, e.g., being a constant that is equal across all types as in \citealt{nowak+etal:2002}). For this purpose, we turn to a different strand of research in cultural evolution: iterated learning.

@@

\subsection{Learnability}\label{sec:learnability}
Iterated learning is a process in which the behavior of one individual serves as learning input for another. This learner, upon acquisition of this behavior, then goes on to produce behavior that serves as input for a new learner. This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language. The latter, now a parent, goes on to produce data for a new generation of learners. Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated process of Bayesian inference in which learners combine the likelihood of a type producing the received learning input with prior inductive biases.

Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for recent surveys). Importantly, experimental and mathematical results have been taken to suggest that the outcome of this process reflects learners' inductive biases. In a Bayesian setting these biases can be codified in a prior $P \in \Delta(T)$, which can be thought of as the amount of data a learner would require in order to adopt a language (cf. \citealt[450]{griffiths+kalish:2007}). Or, in our case, a combination of a lexicon and its use. The extent of the prior's influence has been shown to heavily depend on the learning strategy assumed to underly the inference process. On the one hand, early simulation results suggested that weak biases could be magnified by exposing learners to only small data samples (e.g. in \citealt{brighton:2002}). On the other, \citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated learning converged to the prior, i.e., the resulting distribution over languages corresponds to the learners' prior distribution and is not influenced by the amount of input given to them. This difference in predictions can be traced back to differences in the selection of hypotheses from the posterior. Griffith \& Kalish's convergence to the prior holds for learners that sample from the posterior. More deterministic strategies such as the adoption of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we parametrize the posterior by $l \geq 1$. In doing so, we obtain a range of learning strategies between posterior sampling and MAP. When $l = 1$ learners sample from the posterior and the learners propensity to maximize the posterior grows as $l$ increases.

@@

%To summarize, the parametrized posterior, $P(d|t)^l$, is obtained from combining the prior $P(t)$ and the likelihood $P(d|t)$. Since types are combinations of a lexicon and its use, the latter can be computed straightforwardly from a type's production behavior. 
We combine the above with the replicator dynamics by codifying iterated learning in a transition matrix $Q$, where $Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$. This quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data:

@@

\subsection{Model summary}
Drawing from past research, we argued that expressivity and learnability are central to the cultural evolution of language. We propose these components to be modelled respectively as communicative efficiency-relative replication and (iterated) Bayesian learning. Taken together their interaction is described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:

@@

\section{Scalar implicatures}\label{sec:si-case-study}
%
Scalar implicatures are a particularly well-studied type of conventional pragmatic inferences. They are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order. For instance, {\em some} is entailed by {\em all}. If it were true that `All students came to class', it would also be true that `Some students came to class'. However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey. Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold because the speaker used a weaker alternative instead. In this way, `Some students came to class' is strengthened to convey that some but not all students came to class. A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.

This kind of strengthening is captured by the linguistic behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about a speaker's use of a message involving a weak scalar alternative will associate it more strongly with upper-bounded states than with ones in which stronger alternatives hold because these alternatives already unambiguously convey the latter states. Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly.

@@

We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions. The first is that their truth-conditional compatibility with stronger expressions endows them with a broader range of applicability by allowing them to occur in contexts in which their upper-bounded reading is absent. This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant. For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty. This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: Should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions, e.g., by saying {\em some but not all} explicitly. Importantly, although morphosyntactic disambiguation may be dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer. In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so. Compare this with a hypothetical language that lexicalizes two expressions for each weak scalar expression -- one with and one lacking an upper-bound. We see four conditions along this functionalist explanation that may pressure languages for English-like semantics over this alternative. First, contextual cues are very reliable. Second, morphosyntactic disambiguation is seldom necessary. Third, morphosyntactic disambiguation is only marginally dispreferred. Fourth, larger lexica are costly. Overall, neither condition seems convincing as a pivotal explanatory device for such a widespread phenomenon. The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication. It is likely that human language users are very good at retrieving cues from context, but to stipulate that they are so good as to undercut the benefit of safe communication provided by this hypothetical alternative strikes us as too strong of an assumption.  As for the third and fourth condition, these seem mostly like technical solutions without a proper empirical basis.

Instead, the systematicity and typological spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984}, \citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be at play. In what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences may be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it. This difference is reflected in a learning bias towards more compressed lexical representation, i.e., in a preference of learners for simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
